The conserved genetic program of male germ cells uncovers ancient regulators of human spermatogenesis

Version 1

Abstract

Summary

Male germ cells share a common origin across animal species, therefore they likely retain a conserved genetic program that defines their cellular identity. However, the unique evolutionary dynamics of male germ cells coupled with their widespread leaky transcription pose significant obstacles to the identification of the core spermatogenic program. Through network analysis of the spermatocyte transcriptome of vertebrate and invertebrate species, we describe the conserved evolutionary origin of metazoan male germ cells at the molecular level. We estimate the average functional requirement of a metazoan male germ cell to correspond to the expression of approximately 10,000 protein-coding genes, a third of which defines a genetic scaffold of deeply conserved genes that has been retained throughout evolution. Such scaffold contains a set of 79 functional associations between 104 gene expression regulators that represent a core component of the conserved genetic program of metazoan spermatogenesis. By genetically interfering with the acquisition and maintenance of male germ cell identity, we uncover 161 previously unknown spermatogenesis genes and three new potential genetic causes of human infertility. These findings emphasize the importance of evolutionary history on human reproductive disease and establish a cross-species analytical pipeline that can be repurposed to other cell types and pathologies.

Main Text

Understanding what defines the uniqueness of a given cell type out of the 843 predicted cellular fates in the human body is a complex and fascinating problem1. Through Conrad Waddington’s foundational work, we have come to appreciate that developmental trajectories ultimately dictate cell type identity via the establishment of specific transcriptional programs2. The fact that transcriptomes tend to cluster by tissue type rather than by species3 clearly indicates that gene expression identity can be maintained across many million years of evolutionary divergence. This echoes the modular nature of eukaryotic biological processes, whose intervening macromolecular complexes are typically built by the addition of younger components to a core block of ancient subunits4.

The emergence of germ cells is considered one of the first cell type specializations in metazoan history5. Since the capability to undergo both sexual reproduction and gametogenesis were already present in the unicellular ancestor of all Metazoa6, the split between germ line and soma presumably provided early multicellular organisms increased robustness against mutations while minimizing genetic conflict between different cell lineages7. Yet, reproductive protein genes are some of the most rapidly-evolving across the tree of life8, leading to substantial variation in male reproductive tissues even between closely-related species9,10. At the molecular level, this variation can be mainly traced back to rapid evolutionary changes in gene expression in the testis11 - likely facilitated by widespread transcriptional leakage in male germ cells12 - and to the preferential emergence of new genes in this organ13, where they often acquire functions in spermatogenesis14. Although the rapid divergence of genes directly involved in reproduction has traditionally been interpreted from an adaptationist stance, relaxed selection and drift have been recently proposed to account for this pattern15. Central to this debate is the often-overlooked contribution of old genes for germ cell development and function.

We have recently observed, in a wide range of plant species, a substantial contribution of old genes to the pollen transcriptome, suggestive of an ancient transcriptional program common to plant male gametes16. Such observation finds parallel in the concept of a metazoan germ line multipotency program17, and is supported by transcriptional similarities between equivalent spermatogenic cell types across mammalian species18–20. This led us to explore the possibility that, despite the rapid evolution of reproductive protein genes, the functional basis of metazoan male germ cell identity relies on an old, evolutionarily-conserved genetic program that can provide relevant insight on spermatogenesis and on human infertility.

To test this hypothesis, we devised an interdisciplinary research platform based on four combined approaches (Fig. 1a). First, we determined, through phylostratigraphy, the age of the gene expression program of male germ cells from three evolutionarily-distant metazoan species: humans (Homo sapiens), mice (Mus musculus) and fruit flies (Drosophila melanogaster). Then we used network science to infer the significance of deeply conserved genes within the context of the complex male germ line transcriptome. Subsequently, through developmental biology (in vivo RNAi in fruit fly testes), we defined the role of a key subset of the conserved germ cell transcriptome in male reproductive fitness. Finally, we merged this information with clinical genetics to identify new potential causes of human infertility. Overall, we show that deeply conserved genes play a prominent role in male germ cell regulation and that the disruption of this ancient genetic program leads to human reproductive disease.

The male germ cell transcriptome has an old evolutionary origin.

a- Overview of the experimental strategy. b- The diversity of the male germ cell transcriptome substantially depends on lowly-expressed genes. Three representative somatic cell types are included for comparison. TPM: transcripts per million; see Sup. Table 1 for information on RNA-Seq datasets. c- Clade tree for mapping the time of origin of genes in the three selected species: human (Primata), mouse (Rodentia), and fruit fly (Diptera). Genes assigned to phylostrata 1-5 are common to all metazoan species. Mya: million years ago; see Sup. Fig. 1 for the list of representative species of each clade and number of genes in each phylostratum. d- The majority of genes expressed by male germ cells are common to all Metazoa (phylostrata 1-5, green outline). This fraction is similar to that found in representative somatic cell types of each selected species. Minimum average expression cut-off: TPM >1. ns-no significant difference (p >0.3472; Mann–Whitney U test). Sg.: Spermatogonia, Sc.: Spermatocytes, St.: Spermatids, En.: Enterocytes, Ne.: Neurons and Ms.: Muscle. e- Post-meiotic male germ cells have younger transcriptomes than meiotic and pre-meiotic cells. Transcriptome age indices (TAIs) are split between the different phylostrata. f- The spermatocyte transcriptome forms a large, structured network based on protein-protein interaction (PPI) data. Graphs represent the largest connected component of all spermatocyte-expressed genes (minimum average expression cut-off: TPM >1) according to STRING functional association scores. Gene conservation (across all Metazoa) was defined based on eggNOG orthogroups. Networks were filtered to only include edges with combined scores ≥0.5 (see Sup. Fig. 3). g- Spermatocyte PPI networks contain a substantial number of conserved genes. h- Conserved genes (red) are more connected than non-conserved genes (blue) in both germ cell (spermatocyte) and somatic cell (enterocyte) PPI networks ****p <0.0001 (Kolmogorov-Smirnov test). i- Machine-learning algorithms reliably predict the evolutionary conservation of spermatocyte-expressed genes based solely on PPI network features. Values correspond to AUC (area under the curve) scores. “Coin toss” corresponds to a random classification. Four-fold cross-validation results are shown. ROC: receiver operating characteristic curves; SVM: support-vector machine; see Sup. Fig. 5 for precision and recall curves.

A more conservative estimate of the complexity of the male germ cell transcriptome

Male germ cell development is divided into three conserved stages21. The first is the pre-meiotic stage and corresponds to the mitotic expansion of committed precursors (spermatogonia). Meiosis defines the second stage, with the newly-differentiated spermatocytes undergoing reductive division. In the third stage, the post-meiotic cells (spermatids) embark on a cytodifferentiation program that culminates with the formation of mature male gametes.

To understand to what extent male germ cell transcription quantitatively differs from that of somatic lineages, we collected previously published high-quality RNA-Seq datasets from pre-meiotic, meiotic and post-meiotic germ cell populations, and compared them with representative somatic cell types of the primary embryonic layers: neurons (ectoderm), muscle (mesoderm) and enterocytes (endoderm; Fig. 1b and Sup. Table 1). We centered our analysis on three evolutionarily-distant gonochoric species with excellent genome annotations - humans, mice and fruit flies (the last two also being well-established animal models for the functional study of human spermatogenesis) - and observed that, as previously reported12,22, male germ cells have a generally more diversified transcriptome than their somatic counterparts (as measured by the percentage of the entire genome each cell type expresses). Yet, this increased complexity was mainly restricted to extremely permissive minimum average expression cut-offs ranging from >0.01 to >0.5 transcripts per million (TPM), suggestive of a significant degree of transcriptional noise or other previously hypothesized translation-independent roles for lowly-expressed transcripts23. Indeed, the use of a more suitable threshold for protein-coding transcripts (TPM >1)24 revealed that, on average, metazoan male germ cells tend to express approximately 10,000 genes (9,468 ± 498 to 12,195 ± 1,144, depending on the species), corresponding to half of the entire protein-coding genome (48.0 - 67.9%; Fig. 1b and Sup. Table 1). In both humans and mice, the latter range is comparable to that observed in somatic cell types such as neurons and enterocytes (53.4 - 53.7% and 37.3 - 45.6% of the genome, respectively). Only in fruit flies, a species with a smaller genome (13,947 protein-coding genes vs. 22,802 and 22,287 in humans and mice, respectively), the diversity of the male germ cell transcriptome diverged substantially from that found in representative somatic tissues (with an average of 67.9 ± 3.6% of the genome expressed in germ cells vs. 33.3 ± 12.5% in the soma). It should be noted that this percentage may be higher in fruit flies due to the presence of transcripts of somatic origin in the germ cell dataset (see Sup. Table 1). Collectively, these observations suggest that the average functional requirement of a metazoan male germ cell corresponds to the expression of approximately 10,000 protein-coding genes. Such a number brings a significant but largely overlooked functional constraint to the general notion that almost all protein-coding genes are transcribed in the male germ line.

The male germ cell transcriptome has an old evolutionary origin

We next assessed the contribution of deeply conserved genes to the male germ cell transcriptome. For this we used phylostratigraphy, a technique that determines the evolutionary age of groups of homologous genes (defined as orthogroups in our analysis, including both orthologs and paralogs) by mapping their last common ancestor in a species tree25. We assembled our tree based on the proteomes of 25 phylogenetically-representative eukaryotic species, and assigned each orthogroup to the oldest phylogenetic node (phylostratum) it could be traced back to (Fig. 1c and Sup. Fig. 1). This way, phylostrata ranked 1 to 5 were the oldest and contained orthogroups common to all Metazoa, while phylostratum 16 contained the youngest, species-specific orthogroups. A total of 113,757 orthogroups were identified in the 25 representative eukaryotes, 85.5% of which (97,270) were species-specific when adding all phylostrata. By using TPM >1 as minimum expression cut-off (to minimize the effects of widespread, low-level transcription in the male germ line), we observed that 48.0 - 67.9% of the entire protein-coding genome was expressed in male germ cells, depending on the species. Despite the rapid divergence typically associated with reproduction-related genes, we found that the majority (65.2 to 70.3%) of genes expressed in male germ cells mapped to the oldest-ranking phylostrata containing orthogroups common to all Metazoa, henceforth referred to as deeply conserved genes (Fig. 1d). Indeed, in all three tested species this percentage was not significantly different from that recorded in the transcriptome of the somatic cell types (also at a threshold of TPM >1). The fraction of deeply conserved genes that were ubiquitously expressed (proxy to their involvement in cellular housekeeping processes) varied from 49.7 to 63.4% (Sup. Fig. 2a-b). This strongly suggests that almost half of all deeply conserved genes expressed by male germ cells are involved in more specific roles than just the maintenance of basal cellular functions.

By summing the products of the age of all expressed genes and their expression levels at a given developmental stage (a metric known as the transcriptome age index - TAI26) we determined the transcriptome age of the different male germ cell stages (Fig. 1e). As recently reported in mammalian spermatogenesis20, meiotic and pre-meiotic cells across all three tested species had lower TAIs than post-meiotic cells, indicative of older transcriptomes. This trend was less obvious in the fruit fly as in this species post-meiotic transcription is largely residual27. Collectively, we observed that, both in vertebrates and invertebrates, the male germ cell transcriptome has an old evolutionary origin that is tempered by the increased expression of younger genes at later developmental stages. Our data, indicative of a core spermatogenic program spanning more than 600 million years of evolution, significantly expand the breadth of previous observations in the mammalian lineage18–20, while emphasizing the potential relevance of using conserved expression between distant animal species to gain insight into the molecular basis of human infertility.

Deeply conserved genes are central components of the male germ cell transcriptome

We next addressed the possible significance of the abundant expression of deeply conserved genes in the male germ line. For this we focused on meiosis, a fundamental process conserved through eukaryotes. Besides the existence of an ancient meiotic toolkit required for recombination and reduction division across sexually-reproducing species28, the mitosis-to-meiosis transition in the male germ line is also home to a transcriptional burst associated with the acquisition of male germ cell identity29. The transcripts produced in this burst are mainly required for post-meiotic development, with their translation being shifted towards the later spermatogenic stages30. Accordingly, when analyzing the expression pattern of all human and mouse genes annotated to the “spermatid differentiation” Gene Ontology (GO) term (GO:0048515), we observed that only a small fraction of these was expressed primarily/exclusively after meiosis [5.7% (13 out of 228) and 8.7% (25 out of 289) in humans and mice, respectively]. Similar to previous reports in fruit flies27,29,31, most of the mammalian spermatid differentiation genes were found to be expressed already at the meiotic stage, with 82.9% (189 out of 228, in humans) and 77.9% (225 out of 289, in mice) being expressed in spermatocytes or in spermatocytes and spermatogonia. These results show that the meiotic transcriptome provides a particularly suitable entry-point into the genetic basis of the core spermatogenic program. Hence, we assembled protein-protein interaction (PPI) networks based on the transcriptome data of human, mouse, and fruit fly spermatocytes, where nodes represent all expressed genes and edge weights indicate the probability of the connected genes contributing to a specific function according to the STRING database32 (Fig. 1f). The structure of these networks reflects the multiple genetic interdependencies responsible for cellular function, as illustrated by the characteristic clustering of functionally-related genes into topologically-defined modules33. Network edges were filtered to remove low confidence interactions (those with a combined confidence score <0.5, Sup. Fig. 3a-c), and, as before, only genes expressed at TPM >1 were included to minimize the confounding effects of widespread low-level transcription. The resulting networks contained between 7,961 and 11,322 genes, depending on the species, and an average of 301,036 ± 66,416 edges.

Consistent with phylostratigraphy, the spermatocyte PPI networks included a substantial fraction of genes conserved across all Metazoa (45.1 to 58.1% of all genes; Fig. 1g). As starting point to our analyses, we set out to determine to what extent the previously reported increased connectivity of conserved genes in PPI networks34,35 is maintained in a cell type characterized by rapid rates of evolution. For comparison purposes, we assembled, as before, PPI networks based on the transcriptome data of human, mouse, and fruit fly enterocytes (a somatic cell that, similarly to spermatocytes, is also part of an epithelium). These somatic cell networks, albeit smaller due to lower transcriptome diversity in enterocytes (Fig. 1b), were comparable in size to those of the spermatocytes, containing between 3,310 and 9,837 genes, depending on the species, and an average of 203,296 ± 117,545 edges (Sup. Fig. 3a-d). Similarly to what observed in enterocytes, deeply conserved genes in the spermatocyte PPI networks were significantly more connected than their non-conserved counterparts (higher degree centrality), and their interactors were themselves more connected than those of non-conserved genes (higher page rank, Kolmogorov-Smirnov test for the two analyses; Fig. 1h; Sup. Fig. 4). Although the connectivity properties of the spermatocyte conserved genes were lower than those expressed in enterocytes, they were still salient enough for machine learning classifiers to reliably predict if a gene was conserved just based on spermatocyte network features. More specifically, using a Random Forest classifier, the AUC (area under the receiver operating characteristic curve) score was 0.74, 0.75, and 0.82 in the human, mouse, and fruit fly spermatocyte datasets, respectively (Fig. 1i). Classification performance was equally high when using a different supervised learning algorithm (linear Support Vector Machine) and an alternative performance metric (precision and recall; Sup. Fig. 5).

To address a possible ascertainment bias associated with more available information on conserved genes in the STRING database, we tested to what extent degree centrality and page rank were affected by network rewiring. In this approach, a variable percentage (20 to 100%) of all edges is randomly shuffled across the network, thus diluting any latent biases in the datasets. We observed that both network properties remained higher in conserved genes even when 80% of all edges were rewired in the spermatocyte PPI networks (Sup. Fig. 6). Such result suggests that the increased connectivity of conserved genes is mostly driven by their intrinsic properties, rather than by differences stemming from the amount of source data available. Based on these analyses, we conclude that despite the testis being a rapidly evolving organ and a preferential birthplace for new genes, deeply conserved genes remain central components in the male germ cell transcriptome of evolutionarily-distant species. Accordingly, we decided to explore this increased connectivity as means of gaining insight into the conserved molecular basis of metazoan spermatogenesis.

A core component of the ancient genetic program of spermatogenesis

We have previously shown that it is possible to simplify the complexity of biological networks by removing redundancy and only retaining the most relevant interactions that sustain the dynamics of the system36. To do so, one can extract from a network its distance backbone, i.e., the small subset of metric edges that is sufficient to compute all shortest paths in the original network37. This technique retains the minimum number of required edges to preserve all network nodes, as well as all of the natural hierarchy of connections and community structures38. In the context of biological networks, such distance backbone uncovers the subgraph of gene interactions more likely to convey the key functional characteristics of the gene expression program39.

By extending orthology to the backbone, we have developed a new type of multilayer network: the orthoBackbone. It expands on the concept of interologs40,41, as it consists of only the edges that connect the same pair of orthologs in the network backbone of different species (Fig. 2a). Since this technique excludes all backbone edges (and corresponding nodes) that do not involve the same orthologs across multiple species, it ultimately identifies a subset of conserved genes whose functional interactions at the protein level are part of the distance backbone across evolution. In this regard, the orthoBackbone offers access to what can be considered core features of the genetic program of a given cellular state. Importantly, the use of fruit flies (an animal without male crossing over and recombination) in this inter-species approach significantly reduces the contribution of the ancient meiotic toolkit to the orthoBackbone, thus emphasizing functional interactions associated with spermatogenesis but not with meiosis per se.

Functional analysis of the conserved genetic program of male germ cells.

a- The orthoBackbone methodology. First, the most relevant associations are determined by defining the individual metric backbones (based on shortest paths) of protein-protein interaction (PPI) networks from different species. Of the backbone edges (in green), those connecting the same orthologous genes across the different species are selected as part of the evolutionarily-conserved orthoBackbone (in red, with asterisks). In case of a one-to-many conserved edge relationship, inclusion depends on at least one of the multiple edges being part of the backbone. Letters depict different genes, B’ and B’’ are paralogs, and numbers indicate distances between genes. b- The orthoBackbone represents less than 3% of all functional interactions (edges) in the spermatocyte PPI networks. c- The orthoBackbone connects >70% of all conserved genes expressed in spermatocytes. Gene conservation (across Metazoa) was defined based on eggNOG orthogroups. d- orthoBackbone genes are preferentially involved in gene expression regulation compared with other equally conserved genes. Charts represent the top 5 terms of an unfiltered gene ontology (GO) enrichment analysis for biological processes of the human male germ cell orthoBackbone. False discovery rate ≤0.05; see Sup. Fig. 7 for the expanded GO analyses. e- The male germ cell orthoBackbone reveals a core set of 79 functional interactions between 104 gene expression regulators of spermatogenesis. Solid dots indicate genes with testis specific/enriched expression. Post-transc. reg.: Post-transcriptional regulation; RNA mod.: RNA modification. f- Conserved mitosis-to-meiosis transcriptional burst genes were defined based on their upregulation at mammalian meiotic entry and/or downregulation at meiotic exit. In both cases, genes also had to be expressed in insect spermatogenesis. Green lines link orthologs (920 in fruit flies, 797 in humans and 850 in mice) based on eggNOG orthogroups. Expression level in normalized absolute log(FPKM+1). g- An in vivo RNAi screen in fruit fly testes uncovers the functional requirement of 250 conserved transcriptional burst genes (27.2%) for male reproductive fitness. Silencing of the 920 genes was induced at the mitosis-to-meiosis transition using the bam-GAL4 driver. Color-code for the recorded testicular phenotype as in “h”. Results reflect a total of four independent experiments. Threshold for impaired reproductive fitness (red horizontal line) corresponds to a 75% fertility rate (>2 standard deviations of the mean observed in negative controls). h- Conserved transcriptional burst genes are required for diverse spermatogenic processes. Testicular phenotypes of the 250 hits were defined by phase-contrast microscopy and assigned to five classes based on the earliest manifestation of the phenotype. i- Transcriptional burst genes reveal 161 new, evolutionarily-conserved regulators of spermatogenesis (64.4% of all hits, homologous to 179 and 187 in humans and mice, respectively). Phenotype novelty was defined by lack of previously published evidence of a role in male fertility/spermatogenesis in humans, mice or fruit flies. j- All data acquired in this screen are freely available in the form of an open-access gene browser (Meiotic Navigator).

We observed that the orthoBackbone drastically reduced the size of the spermatocyte PPI networks, retaining only 1.7 to 2.7% of all edges, depending on the species (Fig. 2b). This almost residual fraction of edges nevertheless contained 70% of all deeply conserved genes in the network (Fig. 2c and Sup. Table 2), further illustrating that strong functional relationships between deeply conserved genes are also maintained in the male germ line. GO term enrichment analysis for biological processes revealed that orthoBackbone genes were preferentially involved in gene expression and protein regulation, in contrast with the other (not retained) conserved genes that were mainly associated with cell signaling pathways (Fig. 2d and Sup. Fig. 7). To better understand to what extent the orthoBackbone can provide insight into cell-type specific processes, we determined to the degree of similarity between the spermatocyte and enterocyte orthoBackbones. We noted a modest overlap between both subnetworks, with just 16.2% to 18.0% of common edges depending on the species (against background expectations of 33.6 to 61.5%, based on Jaccard similarity index scores between the PPI networks of both cell types; Sup. Fig. 3d).

Using GO annotations, we further selected spermatocyte orthoBackbone interactions involving gene expression regulators previously linked to spermatogenesis. By doing so we obtained a subset of 79 functional interactions between 104 human genes (out of the total of 3,596 in the orthoBackbone; Fig. 2e). These were ascribed to a wide gamut of regulatory processes and involved noteworthy spermatogenesis genes such as RFX2 (transcription), CDYL (chromatin remodeling) and BOLL (translation), among others42–44. Of the 104 genes, 34 were testis-specific/enriched while 60 had low tissue specificity, consistent with previous observations that cell identity relies on the integration of cell type-specific genes with more broadly-expressed regulators45. Based on its conservation in the orthoBackbone of evolutionarily-distant species and on the significant spermatogenic role of the intervening genes, we posit that this subnetwork of 79 functional interactions involving 104 genes likely represents a core component of the conserved genetic program of spermatogenesis. This number fits well with previous estimates, based on expression trajectories of 1:1 orthologous genes, of an ancestral program of amniote spermatogenesis constituted by 389 genes20.

Male germ cell identity has a broad functional basis

We next set out to determine the functional consequences of disrupting the spermatocyte orthoBackbone. Since the latter represents a still sizeable interaction subnetwork consisting of more than 3,000 genes, we decided to select candidate genes likely to affect the acquisition and maintenance of male germ cell identity. For this, we again took into account how the mitosis-to-meiosis transcriptional burst activates the expression of spermatogenesis genes - a process that has been likened to that of cellular reprogramming46,47. We reasoned that genes that are upregulated at meiotic entry and/or downregulated at meiotic exit (thus part of the transcriptional burst) likely represent instructive elements for the male germ cell state, hence preferential routes to tamper with this cellular identity.

Through differential gene expression analysis, we identified, in human and mouse spermatocytes, all expressed genes that shared a similar upregulation at meiotic entry and/or downregulation at meiotic exit. Of these 970 mammalian differentially expressed genes (DEGs), we selected as candidates for functional assessment only those that were also expressed in fruit fly germ cells (Fig. 2f). We did not take into account differential expression in the latter species, since the largely residual levels of post-meiotic transcription in fruit fly spermatogenesis thwarts direct comparisons with the mammalian system27. The resulting 920 fruit fly genes (homologous to 797 and 850 in humans and mice, respectively) were silenced specifically at the mitosis-to-meiosis transition by Drosophila in vivo RNAi, using the well-established bam-GAL4 driver48. Of these, a total of 250 (27.2%) were essential for male fertility, as their silencing resulted, upon mating with wildtype females, in egg hatching rates below the cut-off of 75% (>2 standard deviations of the mean observed in negative controls: 91.6 ± 8.5%; Fig. 2g). We observed that 76.0 % of the hits (190 out of 250) were part of the orthoBackbone, compared with 60.0% (402 out of 670) in genes without any obvious effect on male reproductive fitness upon silencing, with this difference being statistically significant (two-sided Fisher’s exact test P<0.0001). Cytological analysis of all 250 genes by testicular phase-contrast microscopy in the RNAi males revealed diverse origins for the infertility phenotype, with the earliest manifestation of cellular defects ranging from the pre-meiotic to the mature gamete stage (Fig. 2h). By exploring publicly-available information (see Methods), we determined that 161 (64.4%) of all hits had never been previously associated with male reproduction in any species (Fig. 2i). Accordingly, these 161 new deeply conserved spermatogenesis genes (homologous to 179 and 187 in humans and mice, respectively) represent both a significant advance in our understanding of the genetic basis of male germ cell development, and a valuable resource to explore from a precision medicine perspective. To facilitate open access to this information, we made all data available in the form of a user-friendly gene browser (https://pages.igc.pt/meionav; Fig. 2j). Overall, by merging our results with previously published data we conclude that at least 43.5% of the mitosis-to-meiosis transcriptional burst (our 250 hits + 150 previously reported male fertility genes that gave a negative result in our assay) is required for a diverse range of germ cell functions. This illustrates the pervasive influence of the mitosis-to-meiosis transition on multiple facets of the spermatogenic program and argues for a broad functional basis underlying male germ cell identity.

An ancient regulator of human spermatogenesis

One of our 161 newly-identified spermatogenesis genes - the Drosophila RING finger protein 113 (dRNF113, a spliceosomal component also known as mdlc49) - emerged as a particularly interesting germ cell regulator from a clinical perspective. More specifically, by analyzing our in-house whole exome database containing sequencing data from 74 cases of human male meiotic arrest, we identified an infertile man harboring a homozygous loss of function (LoF) variant in a human paralog (RNF113B). This frameshift variant c.556_565del;p.(Thr186GlyfsTer119) leads to the abrogation of the protein’s two functional domains (a C3H1-type zinc finger and a RING finger) and to its truncation (Fig. 3a and Sup. Fig. 8a). The identified man (M1911) is part of a consanguineous family of Middle-Eastern ancestry and shared the homozygous state of the RNF113B LoF variant with his equally infertile brother, but not with his fertile heterozygous brother (Sup. Fig. 9a-c and Sup. Information). We excluded the possibility that this genetic variant is an indicator of another, nearby variant that is homozygous in the affected individuals and represents the actual cause of infertility in both men (Sup. Information). Remarkably, the results from the testicular biopsy of M1911 revealed an equivalent meiotic arrest phenotype to that observed in dRNF113-silenced fruit flies. Indeed, spermatocytes were, by far, the predominant cell type in the male gonads of both species: they were the most advanced cell stage observed in 89.0% of all assessed human seminiferous tubules (vs. 9% in controls) and occupied an average of 64.1% of the entire testis area in fruit flies (vs. 12.3% in controls; Fig. 3b-c and Sup. Fig. 8b-c). Early (round) spermatids were practically absent in both species, with cellular debris accumulating in the post-meiotic region of the fruit fly gonad. The similar reproductive impairment phenotypes recorded in both species highlights the often-complex relationship between tissue specificity and functional significance. In insects, dRNF113 is a broadly expressed gene essential for organismal viability49, whose particular spermatogenic function was uncovered by germ cell-specific silencing. Both roles were functionally uncoupled in primates via the existence of two RNF113 copies, with the correctly-spliced form of RNF113B being mainly testis-specific50. The existence of an independent duplication event in the rodent lineage offered us the opportunity to further explore this concept. Compared with its human ortholog (RNF113B), the mouse duplication (Rnf113a2) is broadly expressed, similarly to Rnf113a1, human RNF113A and fruit fly dRNF113. Yet, the generation of a whole-body Rnf113a2 knockout (KO) mouse revealed a comparable phenotype to that of the human RNF113B LoF variant: severe spermatogenic impairment without any overt somatic defects. When compared to wildtype littermates, male homozygous Rnf113a2KO/KO mice were sterile and had visibly smaller testes that were largely devoid of germ cells (SCO: Sertoli cell-only phenotype) except for the rare presence of spermatogonia and spermatocytes (Fig. 3d and Sup. Fig. 8d-f).

Deeply conserved regulators of human spermatogenesis.

a- Similar domain structure of the RNF113 proteins in humans (RNF113B), fruit flies (dRNF113) and mice (Rnf113a2). All contain a C3H1-type zinc finger and a RING finger domain. Numbers indicate amino acid residue position. b- Human RNF113B is required for meiotic progression past the primary spermatocyte stage. Testicular histology of M1911 [RNF113B loss of function (LoF) variant] and of a control sample with normal spermatogenesis. See Sup. Fig. 8b for phenotype quantification. Arrowheads: primary spermatocytes; arrows: spermatids. Scale bars: 100 μm (overview), 50 μm (insets), and 10 μm (meiotic region). c- Silencing fruit fly dRNF113 results in meiotic arrest. Phase-contrast microscopy. See Sup. Fig. 8c for phenotype quantification. Arrowheads: primary spermatocytes; asterisks: early (round) spermatids; arrows: late (elongating) spermatids; sv-seminal vesicle. Scale bars: 50 μm (whole testis) and 20 μm (meiotic region). d- Mouse Rnf113a2 is essential for spermatogenesis. Testes of whole-body homozygous Rnf113a2 knockout mice (Rnf113a2KO/KO) are essentially devoid of germ cells, with the rare occurrence of meiotic and pre-meiotic stages. The testicular histology of a wildtype littermate control (Rnf113a2WT/WT) is presented for comparison (normal spermatogenesis). See Sup. Fig. 8d-g for additional data. Arrowheads: primary spermatocytes; arrows: spermatids. Scale bars: 50 μm (overview) and 20 μm (intratubular region). e- Human RNF113B is predominantly expressed at meiotic entry. Data analyzed from our recently published single cell RNA-Seq atlas of normal spermatogenesis51. Usg.: undifferentiated spermatogonia; Dsg.: differentiated spermatogonia / preleptotene; Lp.: leptotene; Zg.: zygotene; Pc.: pachytene; Dp.: diplotene; M: meiotic divisions; Rsp.: round spermatids; Esp.: elongating spermatids. f- The nuclear levels of the fruit fly dRNF113 protein increase at meiotic entry. Images are maximum projections of the entire nuclear volume. Spermatocytes correspond to late prophase I cells. Dotted lines delimit the nuclear envelope (as assessed by fluorescent wheat germ agglutinin). Scale bar: 5 μm. a.u-arbitrary units. ****p <0.0001 (unpaired t-test). g- RNF113B is required for normal gene expression during human spermatogenesis. Differential gene expression (DGE) analysis of RNA-Seq data obtained from testicular biopsies of M1911 (RNF113B LoF variant, left and right testis) and of three controls with normal spermatogenesis. Down and upregulated genes in blue and red, respectively. orthoBackbone differentially expressed genes (DEGs) are outlined. FC: fold change. FDR: false discovery rate. Edge disruption corresponds to the number of orthoBackbone edges containing at least one DEG. h- dRNF113 regulates gene expression in the fruit fly male gonad. Whole testes samples (in triplicate) in both experimental conditions. i- Network of functional associations between orthoBackbone genes downregulated both in the RNF113B LoF and dRNF113 RNAi. Node size indicates result of the page rank metric in the spermatocyte protein-protein interaction network (measure of the connectivity of interacting genes), and color specifies if the gene has a known role in spermatogenesis (in any species). Testicular phenotype of men affected by variants in HSPA2 and KPNA2 (red nodes) are depicted in “j”. Edge thickness indicates STRING functional association scores and color specifies main source of data for the associations. j- LoF variants in the orthoBackbone genes HSPA2 and KPNA2 are potentially associated with human male infertility. Testicular histology of individuals M2190 and M2098 (HSPA2 and KPNA2 variants, respectively) reveals a complete loss of germ cells (Sertoli cell-only phenotype). Arrowheads: Sertoli cells. Scale bars: 100 μm (overview), 50 μm (insets), and 10 μm (intratubular region).

Strengthening our interest in the RNF113 genes was their possible involvement with male germ cell identity. By analyzing our recently published single-cell RNA-Seq dataset of normal human spermatogenesis51, we observed that RNF113B was predominantly expressed at the mitosis-to-meiosis transcriptional burst, peaking at the diplotene stage of prophase I (Fig. 3e). This mirrored the protein localization pattern of fruit fly dRNF113, characterized by a substantial nuclear accumulation in primary spermatocytes (Fig. 3f). Also in mice, Rnf113a2 was expressed at the transcriptional burst, with an earlier expression in undifferentiated spermatogonia being the likely cause of the slightly more severe histological phenotype of the Rnf113a2KO/KO mice (Sup. Fig. 8g). Consistent with their role in gene expression, the analysis of M1911’s testicular transcriptome via RNA-Seq revealed the deregulation of 21.7% of all expressed genes (Fig. 3g). Notably, the deregulation detected in M1911’s testicular transcriptome had a clear impact on the spermatocyte orthoBackbone, with 30.0% of its network edges being disrupted (i.e., containing at least one deregulated gene). A similar effect was patent in the dRNF113 RNAi, with 20.3% of the fruit fly testicular transcriptome being deregulated, resulting in the disruption of 26.3% of the orthoBackbone (Fig. 3h). For comparison, the silencing of Prp19, another transcriptional burst gene and spliceosomal component that RNF113 proteins associate to52, had a lower effect on the orthoBackbone with 15.2% of disrupted edges (significantly lower than in the dRNF113 RNAi; two-sided Fisher’s exact test P<0.0001), despite a similar meiotic arrest phenotype (Sup. Fig. 10a-d). Based on the above, we conclude that the RNF113 genes have retained a key role in spermatogenesis for more than 600 million years of evolution. Furthermore, they provided us the means to interfere with a sizable fraction of the male germ cell orthoBackbone in humans and fruit flies - an advantage we next explored from a clinical perspective.

The orthoBackbone as an ancillary tool in clinical genetics

Male infertility is a complex human disease with a poorly-defined genetic component. This contributes to a low causative diagnostic yield (typically just 30%), and to a paucity of clinically-validated genes [currently just 104, in contrast with the more than 700 already associated with other clinical disorders such as intellectual disability53,54]. Since male infertility affects up to 12% of all men55, addressing such knowledge gap is an issue of clear medical importance. To attempt to narrow the gap, we explored the possibility that spermatogenesis is particularly sensitive to disturbances in the male germ cell orthoBackbone. Thus, we harnessed the sizeable effect of RNF113B on this network as means of identifying additional genetic causes of human infertility.

Indeed, we noted that 38.0% (30 out of 79) of the functional interactions that define our previously identified core component of the conserved genetic program of spermatogenesis were disrupted in M1911 (i.e., these interactions contained at least one differentially expressed gene; Sup. Fig. 11). In particular, we observed that 19.2% (20 out of 104) of this gene subset was found to be downregulated, compared with 10.0% (1,685 out of 16,824) of downregulated genes in the testicular transcriptome of M1911, with this difference being statistically different (two-sided Fisher’s exact test P=0.0046; Fig. 3g). Therefore, we posited that the similar spermatogenic impairment recorded in the human RNF113B LoF variant and on the fruit fly dRNF113 RNAi ultimately reflected the downregulation, in both species, of a common set of orthoBackbone genes.

By defining the overlap between differentially expressed orthoBackbone genes in the testicular transcriptome of M1911 and of the dRNF113 RNAi, we identified 61 conserved human genes that were similarly downregulated in both species. These formed a connected network (based on STRING data) suggestive of their functional co-involvement in related biological processes (Fig. 3i). Of the 61 genes, 27 had already been linked to male germ cell defects in different species: four were clinically-validated male infertility genes (CDC14A, CFAP91, DNAI1 and DNAI2), and the other 23 had been previously reported in animal models (see Methods). The fact that among the latter was BOLL, one of the oldest known metazoan gametogenesis genes44, was a particularly noteworthy observation. We next tested if these 27 genes could be used to identify new genetic causes of human infertility. For this we analyzed whole-exome sequencing data of 1,021 azoospermic men from the MERGE cohort56. Filtering these exomes for LoF variants in the 27 selected orthoBackbone genes revealed two new human male infertility candidate genes: HSPA2 and KPNA2.

HSPA2 - heat shock protein family A member 2 - is a molecular chaperone of the highly conserved 70-kDa heat shock protein (HSP70) gene family. HSP70 members are involved in cellular proteostasis from bacteria to human, with the mouse and fruit fly HSPA2 homologs (Hspa2 and Hsc70-1 to Hsc70-5, respectively) being required for meiotic progression past the primary spermatocyte stage57,58. Indeed, we observed that the silencing of Hsc70-1 in the fruit fly testis, also with the bam-GAL4 driver, resulted in an equivalent meiotic arrest to that of dRNF113, strongly suggesting that the misexpression of this chaperone is a central element of the latter phenotype (Sup. Fig. 12a-b). Two different heterozygous LoF variants were detected in our male infertility cohort, with this gene having a predicted autosomal-dominant inheritance (Sup. Fig. 12c-f, Sup. Information and Methods). The first variant, detected in individual M1678, is the early stop-gain variant c.175C>T;p.(Gln59Ter) that truncates more than 90% of the protein, likely leading to nonsense-mediated decay. There is only one individual listed in the gnomAD database with this variant (minor allele frequency = 0.0004%). The second, identified in individual M2190, is the frameshift variant c.1806dup;p.(Glu603ArgfsTer81) which affects the distal end of the protein’s nucleotide-binding domain. This variant is not listed in the gnomAD database. Histopathological analysis of M2190’s testicular tissue revealed a SCO phenotype in 259 assessed seminiferous tubules, a likely aggravation of the extensive cell death accompanied by tubule vacuolization reported in Hspa2 mutant mice59 (Fig. 3j).

KPNA2 - karyopherin subunit alpha 2 - is a nuclear importin that functions as an adapter protein in nucleocytoplasmic transport. Its mouse ortholog (Kpna2) is required for the nuclear accumulation of Hop260, a conserved regulator of meiotic progression from yeast to mammals61,62. The main function of Hop2 is to repair meiotic DNA double-strand breaks (DSBs), with the corresponding mouse mutant being characterized by a primary spermatocyte arrest coupled to extensive cell death. Since male fruit flies dispense with meiotic DSBs63, the inclusion of KPNA2 in the male germ cell orthoBackbone seems counter-intuitive. Yet, Drosophila importin alpha 2 (dKPNA2, also known as Pendulin) is required for post-meiotic development64, with its silencing specifically at meiotic entry being associated with aborted spermiogenesis (Sup. Fig. 13a-b). This post-meiotic function might also be present in its mammalian orthologs, as suggested by the nuclear localization of Kpna2 in elongating mouse spermatids60. We detected two heterozygous KPNA2 LoF variants in our cohort of infertile men, with this gene also having a predicted autosomal-dominant inheritance (Sup. Fig. 13c-g, Sup. Information and Methods). The splice-site variant (c.667-2A>G;p.?) in individual M1645 is predicted to disrupt the correct splicing of intron 6 due to the loss of an acceptor site. Segregation analysis revealed the de novo occurrence of this variant, strongly supporting both its pathogenicity and our recent results pointing to the important contribution of heterozygous de novo mutations to male infertility65,66. The other KPNA2 LoF variant, in individual M2098, is the frameshift variant c.1158_1161del;p.(Ser387ArgfsTer14) which affects 27% of the protein, including its minor nuclear localization signal binding site. Both KPNA2 variants are not listed in the gnomAD database. The available testicular histopathology report for the latter individual lists a SCO phenotype as the cause of the azoospermia, again suggestive of the possible deterioration of an initially meiotic phenotype (Fig. 3j).

Based on three in silico prediction tools (Sup. Information and Methods), both HSPA2 and KPNA2 can represent haploinsufficient genes and/or have autosomal dominant inheritance, with the evidence being stronger for KPNA2. Together with the identified de novo occurrence of one of its LoF variants, this makes KPNA2 a strong novel autosomal-dominant candidate gene while more data is still needed before drawing similar conclusions on HSPA2.

In summary, by experimentally disrupting the male germ cell orthoBackbone across evolutionarily-distant species, we were able to uncover two new candidate genes for human infertility (HSPA2 and KPNA2) in addition to RNF113B, affecting a combined total of 5 individuals. This successful merger between basic and clinical research highlights the advantages of comparative biology as means of dampening inter-species differences in reproductive physiology, while providing a conceptual framework for a more efficient prioritization of clinically-relevant genetic variants in human reproductive disease.

Discussion

We propose that the transcriptional identity of metazoan male germ cells is built around a relatively small network of deeply conserved gene interactions with an overarching functional impact. These observations may provide a unifying genetic basis for the deep conservation of fundamental germ cell biological processes such as sperm motility and gamete fusion67,68. The existence of a shared genetic identity in metazoan spermatocytes can be regarded as a ramification of an ancestral multipotency program already present in germ line precursor cells17. Indeed, even in species with divergent germ line segregation strategies, conserved functional interactions at the post-translational level are required for primordial germ cell specification69,70. It should be noted that our conclusions were based on a small fraction of the diversity of all metazoan species, as a result of data availability limitations and our focus on well-characterized gonochoric animal models of human spermatogenic impairment. The advent of more comprehensive gene expression datasets from less-studied organisms will surely provide a more encompassing context to cross-species comparisons of this nature.

The actual benefit of comparative biology for the identification of new genetic causes of human disease is often a contentious topic71. Central to this debate is the moderate success in translating animal data to the clinical setting, coupled with the fact that the vast majority of human genetic variants are not shared with other species72. By focusing on the deep evolutionary past of human spermatogenesis, we have identified 79 functional interactions that likely represent a core component of the conserved genetic program of male germ cells, and 179 novel functionally-validated candidate genes. The latter have so far uncovered three new genes associated with human male infertility (RNF113B, HSPA2 and KPNA2). The assumed autosomal dominant inheritance of HSPA2 and KPNA2 awaits further confirmation by additional data from independent cohorts and segregation analyses, but stands in line with other well-established male infertility genes such as DMRT1, NR5A1 and SYCP253. Also, this information can potentially be harnessed in other clinically-relevant contexts: the spermatocyte orthoBackbone can be further explored to pinpoint key pathways that may facilitate the reconstitution of meiotic entry in vitro, as well as to maximize the effectiveness of exome sequencing approaches in infertile men through the inclusion of the orthoBackbone genes in prioritization lists for disease-causing variants.

Taken together, our results emphasize the often-overlooked contribution of evolutionary history to human disease and illustrate how interdisciplinary research can significantly expand our knowledge of fundamental cellular processes. Accordingly, all the code required for repurposing our analytical pipeline to other cell types and pathologies is available as an open-access resource at https://github.com/rionbr/meionav. These resources will likely contribute to a renewed appreciation of comparative biology in the medical field.

Supporting information

eLife Assessment

This important study reports the deep evolutionary conservation of a core genetic program regulating spermatogenesis in flies, mice, and humans. The data presented are supportive of the main conclusion and generally convincing. This work will be of interest to evolutionary and reproductive biologists.

Reviewer #1 (Public Review):

Summary:

By combining an analysis of the evolutionary age of the genes expressed in male germ cells, a study of genes associated with spermatocyte protein-protein interaction networks and functional experiments in Drosophila, Brattig-Correia and colleagues provide evidence for an ancient origin of the genetic program underlying metazoan spermatogenesis. This leads to identifying a relatively small core set of functional interactions between deeply conserved gene expression regulators, whose impairment is then shown to be associated with cases of human male infertility.

Strengths:

In my opinion, the work is important for three different reasons. First, it shows that, even though reproductive genes can evolve rapidly and male germ cells display a significant level of transcriptional noise, it is still possible to obtain convincing evidence that a conserved core of functionally interacting genes lies at the basis of the male germ transcriptome. Second, it reports an experimental strategy that could also be applied to gene networks involved in different biological problems. Third, the authors make a compelling case that, due to its effects on human spermatogenesis, disruption of the male germ cell orthoBackbone can be exploited to identify new genetic causes of infertility.

Weaknesses:

The main strength of the general approach followed by the authors is, inevitably, also a weakness. This is because a study rooted in comparative biology is unlikely to identify newly emerged genes that may adopt key roles in processes such as species-specific gamete recognition. Additionally, using a TPM >1 threshold for protein-coding transcripts may exclude genes, such as those encoding proteins required for gamete fusion, which are thought to be expressed at a very low level. Although these considerations raise the possibility that the chosen approach may miss information that, depending on the species, could be potentially highly functionally important, this by no means reduces its value in identifying genes belonging to the conserved genetic program of spermatogenesis.

Reviewer #2 (Public Review):

Summary:

This is a tour de force study that aims to understand the genetic basis of male germ cell development across three animal species (human, mouse, and flies) by performing a genetic program conservation analysis (using phylostratigraphy and network science) with a special emphasis on genes that peak or decline during mitosis-to-meiosis. This analysis, in agreement with previous findings, reveals that several genes active during and before meiosis are deeply conserved across species, suggesting ancient regulatory mechanisms. To identify critical genes in germ cell development, the investigators integrated clinical genetics data, performing gene knockdown and knockout experiments in both mice and flies. Specifically, over 900 conserved genes were investigated in flies, with three of these genes further studied in mice. Of the 900 genes in flies, ~250 RNAi knockdowns had fertility phenotypes. The fertility phenotypes for the fly data can be viewed using the following browser link: https://pages.igc.pt/meionav. The scope of target gene validation is impressive. Below are a few minor comments.

(1) In Supplemental Figure 2, it is notable that enterocyte transcriptomes are predominantly composed of younger genes, contrasting with the genetic age profile observed in brain and muscle cells. This difference is an intriguing observation and it would be curious to hear the author's comments.

(2) Regarding the document, the figures provided only include supplemental data; none of the main text figures are in the full PDF.

(3) Lastly, it would be great to section and stain mouse testis to classify the different stages of arrest during meiosis for each of the mouse mutants in order to compare more precisely to flies.

This paper serves as a vital resource, emphasizing that only through the analysis of hundreds of genes can we prioritize essential genes for germ cell development. its remarkable that about 60% of conserved genes have no apparent phenotype during germ cell development.

Strengths:

The high-throughput screening was conducted on a conserved network of 920 genes expressed during the mitosis-to-meiosis transition. Approximately 250 of these genes were associated with fertility phenotypes. Notably, mutations in 5 of the 250 genes have been identified in human male infertility patients. Furthermore, 3 of these genes were modeled in mice, where they were also linked to infertility. This study establishes a crucial groundwork for future investigations into germ cell development genes, aiming to delineate their essential roles and functions.

Weaknesses:

The fertility phenotyping in this study is limited, yet dissecting the mechanistic roles of these proteins falls beyond its scope. Nevertheless, this work serves as an invaluable resource for further exploration of specific genes of interest.

Author response:

The following is the authors’ response to the original reviews.

eLife assessment:

This important study reports the deep evolutionary conservation of a core genetic program regulating spermatogenesis in flies, mice, and humans. The data presented are supportive of the main conclusion and generally convincing. This work will be of interest to evolutionary and reproductive biologists.

The Authors would like to thank the Senior Editor and the two Reviewers for their positive assessment of our work, as well as for the helpful suggestions. Collectively, these suggestions provided insight that was instrumental in shaping the final version of the manuscript (see below for our point-by-point comments). The Authors believe that the refinements introduced to the final document clearly translate into an improved version of our work. Hence, we would like to thank all those involved in the peer review process for their encouraging words and constructive criticism.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

By combining an analysis of the evolutionary age of the genes expressed in male germ cells, a study of genes associated with spermatocyte protein-protein interaction networks and functional experiments in Drosophila, Brattig-Correia and colleagues provide evidence for an ancient origin of the genetic program underlying metazoan spermatogenesis. This leads to identifying a relatively small core set of functional interactions between deeply conserved gene expression regulators, whose impairment is then shown to be associated with cases of human male infertility.

Strengths:

In my opinion, the work is important for three different reasons. First, it shows that, even though reproductive genes can evolve rapidly and male germ cells display a significant level of transcriptional noise, it is still possible to obtain convincing evidence that a conserved core of functionally interacting genes lies at the basis of the male germ transcriptome. Second, it reports an experimental strategy that could also be applied to gene networks involved in different biological problems. Third, the authors make a compelling case that, due to its effects on human spermatogenesis, disruption of the male germ cell orthoBackbone can be exploited to identify new genetic causes of infertility.

We thank the Reviewer for their positive assessment. Indeed, it was our main objective to convincingly demonstrate these three points.

Weaknesses:

The main strength of the general approach followed by the authors is, inevitably, also a weakness. This is because a study rooted in comparative biology is unlikely to identify newly emerged genes that may adopt key roles in processes such as species-specific gamete recognition. Additionally, using a TPM >1 threshold for protein-coding transcripts may exclude genes, such as those encoding proteins required for gamete fusion, which are thought to be expressed at a very low level. Although these considerations raise the possibility that the chosen approach may miss information that, depending on the species, could be potentially highly functionally important, this by no means reduces its value in identifying genes belonging to the conserved genetic program of spermatogenesis.

The Authors acknowledge the points raised by the Reviewer as inevitable trade-offs of the focus of our study (to uncover the deeply conserved genetic basis of spermatogenesis). Certainly, our pipeline could, in the future, be adapted to look for newly emerged genes or to employ different minimum expression cut-offs. To this end, we made all computational data and custom scripts easily available to the community. We would, nevertheless, kindly emphasize the challenge associated with the use of less restrictive TPM cut-offs, given the substantial level of transcriptional noise associated with this cell type. An abridged version of this discussion can be found in lines 512-515 of the manuscript.

Reviewer #2 (Public Review):

Summary:

This is a tour de force study that aims to understand the genetic basis of male germ cell development across three animal species (human, mouse, and flies) by performing a genetic program conservation analysis (using phylostratigraphy and network science) with a special emphasis on genes that peak or decline during mitosis-to-meiosis. This analysis, in agreement with previous findings, reveals that several genes active during and before meiosis are deeply conserved across species, suggesting ancient regulatory mechanisms. To identify critical genes in germ cell development, the investigators integrated clinical genetics data, performing gene knockdown and knockout experiments in both mice and flies. Specifically, over 900 conserved genes were investigated in flies, with three of these genes further studied in mice. Of the 900 genes in flies, ~250 RNAi knockdowns had fertility phenotypes. The fertility phenotypes for the fly data can be viewed using the following browser link:https://pages.igc.pt/meionav. The scope of target gene validation is impressive. Below are a few minor comments.

We thank the Reviewer for their positive appraisal of our work.

(1) In Supplemental Figure 2, it is notable that enterocyte transcriptomes are predominantly composed of younger genes, contrasting with the genetic age profile observed in brain and muscle cells. This difference is an intriguing observation and it would be curious to hear the author's comments.

Indeed, this is an intriguing observation for which we can only provide a speculative answer. Enterocytes are specialized to absorb nutrients, hence their genetic program is finely tuned to maximize uptake under specific dietary conditions. In this regard, we can posit that variations in nutrient preference/availability in the course of each species’ evolutionary history (associated with habitat, environmental and/or behavioral changes) may have exerted a selective pressure for the emergence of new genes that could provide enterocytes with more efficient uptake capabilities under new circumstances. The application of evolutionary thinking to the rapidly expanding field of nutrigenomics could shed light on this possibility.

(2) Regarding the document, the figures provided only include supplemental data; none of the main text figures are in the full PDF.

We thank the Reviewer for this helpful comment. We will ensure that the three main figures are correctly formatted in the final version of the manuscript.

(3) Lastly, it would be great to section and stain mouse testis to classify the different stages of arrest during meiosis for each of the mouse mutants in order to compare more precisely to flies.

We agree with the Reviewer that adding more mouse data would further improve what can already be considered an extensive body of experimental work. Given the costs associated with the generation of such data (in terms of resources and otherwise), the Authors believe such a study would be best suited to a follow-up manuscript.

This paper serves as a vital resource, emphasizing that only through the analysis of hundreds of genes can we prioritize essential genes for germ cell development. its remarkable that about 60% of conserved genes have no apparent phenotype during germ cell development.

Once again, we thank the Reviewer for their positive assessment of our work. Clarifying the degree of functional redundancy in an essential biological process such as male gametogenesis represents an exciting (and experimentally complex) future challenge.

Strengths:

The high-throughput screening was conducted on a conserved network of 920 genes expressed during the mitosis-to-meiosis transition. Approximately 250 of these genes were associated with fertility phenotypes. Notably, mutations in 5 of the 250 genes have been identified in human male infertility patients. Furthermore, 3 of these genes were modeled in mice, where they were also linked to infertility.

This study establishes a crucial groundwork for future investigations into germ cell development genes, aiming to delineate their essential roles and functions.

The Authors thank the Reviewer for emphasizing the potential usefulness of our results to the community, as that was one of the main motivations behind this project.

Weaknesses:

The fertility phenotyping in this study is limited, yet dissecting the mechanistic roles of these proteins falls beyond its scope. Nevertheless, this work serves as an invaluable resource for further exploration of specific genes of interest.

Please see the previous point.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Although the manuscript already includes a significant amount of data, there are two aspects that the authors may consider exploring:

(1) I understand that the choice of species whose gene expression was analyzed in the study was largely influenced by the quality of the corresponding genome annotations. However, since in evolutionary terms humans and mice are much closer to each other than Drosophila (as also shown in Figure 1c and Supplementary Figure 1), I found the statement "three evolutionarily distant gonochoric species" partially questionable. Have the authors considered adding an additional established animal model, such as for example zebrafish, to provide further coverage of the evolutionary space? Or, alternatively, could a posteriori analysis of the transcriptome of such an additional species be used to cross-validate their findings? The authors touch upon this point in the Discussion, but I wonder if they actually tried something in this direction, or simply decided that the currently available expression data from other organisms was too poor to be used for this purpose.

We thank the Reviewer for bringing up this point, as it echoes one of our main concerns in terms of our approach (as discussed in lines 487-492). Indeed, when we were designing our study, we extensively discussed whether zebrafish and C. elegans datasets should be included, as high-quality expression and phenotypical data were available for both species. We ended up not including them for one main reason: the sexual system of these species deviates from that of humans, mice and fruit flies (all gonochoric species). More specifically, C. elegans are hermaphrodites and although zebrafish is a gonochoric species at the adult stage, they start their lifecycle as juvenile hermaphrodites (they first develop juvenile ovaries that later degenerate into a testis in males). Since it is largely unknown to what extent the transcriptome of male germ cells from these species deviates from the gonochoric program (by retaining oogenesis-related characteristics, for example), we decided to avoid possible confounding effects by excluding the two species. Undoubtedly, as more transcriptomic data from non-model organisms become available, these (and other) questions can be extensively revisited as our pipeline was designed to easily accommodate new data.

(2) Although the use of the STRING database is a sensible choice given the general purpose of this work, in my experience the reliability of its individual interactions can vary significantly. I wonder if the authors have considered exploiting AlphaFold-Multimer as a parallel approach to estimate what proportion of the 79 functional interactions that they identified may reflect direct protein-protein contacts.

We thank the Reviewer for this question and suggestion, as we were also concerned about STRING's reliability for individual interactions. For that reason, we only utilized protein-protein interactions with a STRING combined confidence score ≥0.5 (corresponding to the estimated likelihood of a given association being true), as described in more detail in the "Protein-protein interaction (PPI) network construction" subsection. In addition, to make sure we were not biasing results towards conserved genes (which could arguably be overrepresented in STRING) we pursued a random rewiring test of degree centrality and page rank, as detailed in section "Deeply conserved genes are central components of the male germ cell transcriptome". We very much like the suggestion of using AlphaFold-Multimer to estimate the proportion of direct protein-protein contacts for the 79 core interactions, but given the already quite complex analytical pipeline of the present work, we will leave such analysis for a follow-up study. The final version of the manuscript now contains a reference to such an approach (lines 499-502).

Finally, probably because my primary focus is not on gene regulation, I must say that I found the manuscript somewhat heavy to read. The integration of various data types and analyses, while enriching, also complicates the ability to clearly recall the main conclusions of each result section by the time one reaches the summary at the beginning of the Discussion. Given the relative brevity of the latter, expanding it to both reiterate what these conclusions are and illustrate how all the components converge to support the central message of the study would, in my opinion, benefit a general readership.

We thank the Reviewer for their fresh perspective on our document and for this most welcome suggestion. The final version of the manuscript now includes a longer discussion, containing an initial paragraph (lines 467-479) that summarizes our main findings and how they converge into a coherent body of work.

Additionally, on a minor note, I suggest that the concept of phylostratigraphy be briefly explained when first mentioned in the Introduction, rather than later in the manuscript. This early clarification would aid comprehension for readers unfamiliar with the term.

To safeguard the flow of the manuscript, we have slightly tweaked the introduction section to avoid the use of highly specific terminology (such as phylostratigraphy) this early in the text. We replaced it with “comparison of genome sequences” (line 85). Phylostratigraphy is later explained in full detail in the corresponding section of the manuscript. We thank the Reviewer for this helpful suggestion.

Reviewer #2 (Recommendations For The Authors):

Major concern - the absence of main text figures.

We thank the Reviewer for this helpful comment. We will ensure that the three main figures are correctly formatted in the final version of the manuscript.

Typos throughout - this will need your attention.

The Authors thank the Reviewer for the thorough and attentive assessment of our work. We have carefully revised the text to ensure a pleasant reading experience free of typographical errors.

Version 2

Abstract

Summary

Male germ cells share a common origin across animal species, therefore they likely retain a conserved genetic program that defines their cellular identity. However, the unique evolutionary dynamics of male germ cells coupled with their widespread leaky transcription pose significant obstacles to the identification of the core spermatogenic program. Through network analysis of the spermatocyte transcriptome of vertebrate and invertebrate species, we describe the conserved evolutionary origin of metazoan male germ cells at the molecular level. We estimate the average functional requirement of a metazoan male germ cell to correspond to the expression of approximately 10,000 protein-coding genes, a third of which defines a genetic scaffold of deeply conserved genes that has been retained throughout evolution. Such scaffold contains a set of 79 functional associations between 104 gene expression regulators that represent a core component of the conserved genetic program of metazoan spermatogenesis. By genetically interfering with the acquisition and maintenance of male germ cell identity, we uncover 161 previously unknown spermatogenesis genes and three new potential genetic causes of human infertility. These findings emphasize the importance of evolutionary history on human reproductive disease and establish a cross-species analytical pipeline that can be repurposed to other cell types and pathologies.

Main Text

Understanding what defines the uniqueness of a given cell type out of the 843 predicted cellular fates in the human body is a complex and fascinating problem1. Through Conrad Waddington’s foundational work, we have come to appreciate that developmental trajectories ultimately dictate cell type identity via the establishment of specific transcriptional programs2. The fact that transcriptomes tend to cluster by tissue type rather than by species3 clearly indicates that gene expression identity can be maintained across many million years of evolutionary divergence. This echoes the modular nature of eukaryotic biological processes, whose intervening macromolecular complexes are typically built by the addition of younger components to a core block of ancient subunits4.

The emergence of germ cells is considered one of the first cell type specializations in metazoan history5. Since the capability to undergo both sexual reproduction and gametogenesis were already present in the unicellular ancestor of all Metazoa6, the split between germ line and soma presumably provided early multicellular organisms increased robustness against mutations while minimizing genetic conflict between different cell lineages7. Yet, reproductive protein genes are some of the most rapidly evolving across the tree of life8, leading to substantial variation in male reproductive tissues even between closely related species9,10. At the molecular level, this variation can be mainly traced back to rapid evolutionary changes in gene expression in the testis11 - likely facilitated by widespread transcriptional leakage in male germ cells12 - and to the preferential emergence of new genes in this organ13, where they often acquire functions in spermatogenesis14. Although the rapid divergence of genes directly involved in reproduction has traditionally been interpreted from an adaptationist stance, relaxed selection and drift have been recently proposed to account for this pattern15. Central to this debate is the often-overlooked contribution of old genes for germ cell development and function.

We have recently observed, in a wide range of plant species, a substantial contribution of old genes to the pollen transcriptome, suggestive of an ancient transcriptional program common to plant male gametes16. Such observation finds parallel in the concept of a metazoan germ line multipotency program17, and is supported by transcriptional similarities between equivalent spermatogenic cell types across mammalian species18–20. This led us to explore the possibility that, despite the rapid evolution of reproductive protein genes, the functional basis of metazoan male germ cell identity relies on an old, evolutionarily conserved genetic program that can provide relevant insight into spermatogenesis and human infertility.

To test this hypothesis, we devised an interdisciplinary research platform based on four combined approaches (Fig. 1a). First, we determined, through comparison of genome sequences, the age of the gene expression program of male germ cells from three evolutionarily distant metazoan species: humans (Homo sapiens), mice (Mus musculus) and fruit flies (Drosophila melanogaster). Then, we used network science to infer the significance of deeply conserved genes within the context of the complex male germ line transcriptome. Subsequently, through developmental biology (in vivo RNAi in fruit fly testes), we defined the role of a key subset of the conserved germ cell transcriptome in male reproductive fitness. Finally, we merged this information with clinical genetics to identify new potential causes of human infertility. Overall, we show that deeply conserved genes play a prominent role in male germ cell regulation and that the disruption of this ancient genetic program leads to human reproductive disease.

The male germ cell transcriptome has an old evolutionary origin. a-

Overview of the experimental strategy. b- The diversity of the male germ cell transcriptome substantially depends on lowly-expressed genes. Three representative somatic cell types are included for comparison. TPM: transcripts per million; see Sup. Table 1 for information on RNA-Seq datasets. c- Clade tree for mapping the time of origin of genes in the three selected species: human (Primata), mouse (Rodentia), and fruit fly (Diptera). Genes assigned to phylostrata 1-5 are common to all metazoan species. Mya: million years ago; see Sup. Fig. 1 for the list of representative species of each clade and the number of genes in each phylostratum. d- The majority of genes expressed by male germ cells are common to all Metazoa (phylostrata 1-5, green outline). This fraction is similar to that found in representative somatic cell types of each selected species. Minimum average expression cut-off: TPM >1. ns- no significant difference (p >0.3472; Mann–Whitney U test). Sg.: Spermatogonia, Sc.: Spermatocytes, St.: Spermatids, En.: Enterocytes, Ne.: Neurons and Ms.: Muscle. e- Post-meiotic male germ cells have younger transcriptomes than meiotic and pre-meiotic cells. Transcriptome age indices (TAIs) are split between the different phylostrata. f- The spermatocyte transcriptome forms a large, structured network based on protein-protein interaction (PPI) data. Graphs represent the largest connected component of all spermatocyte-expressed genes (minimum average expression cut-off: TPM >1) according to STRING functional association scores. Gene conservation (across all Metazoa) was defined based on eggNOG orthogroups. Networks were filtered to only include edges with combined scores ≥0.5 (see Sup. Fig. 3). g- Spermatocyte PPI networks contain a substantial number of conserved genes. h- Conserved genes (red) are more connected than non-conserved genes (blue) in both germ cell (spermatocyte) and somatic cell (enterocyte) PPI networks ****p <0.0001 (Kolmogorov-Smirnov test). i- Machine-learning algorithms reliably predict the evolutionary conservation of spermatocyte-expressed genes based solely on PPI network features. Values correspond to AUC (area under the curve) scores. “Coin toss” corresponds to a random classification. Four-fold cross-validation results are shown. ROC: receiver operating characteristic curves; SVM: support-vector machine; see Sup. Fig. 5 for precision and recall curves.

A more conservative estimate of the complexity of the male germ cell transcriptome

Male germ cell development is divided into three conserved stages21. The first is the pre-meiotic stage and corresponds to the mitotic expansion of committed precursors (spermatogonia). Meiosis defines the second stage, with the newly differentiated spermatocytes undergoing reductive division. In the third stage, the post-meiotic cells (spermatids) embark on a cytodifferentiation program that culminates in the formation of mature male gametes.

To understand to what extent male germ cell transcription quantitatively differs from that of somatic lineages, we collected previously published high-quality RNA-Seq datasets from pre- meiotic, meiotic and post-meiotic germ cell populations, and compared them with representative somatic cell types of the primary embryonic layers: neurons (ectoderm), muscle (mesoderm) and enterocytes (endoderm; Fig. 1b and Sup. Table 1). We centered our analysis on three evolutionarily distant gonochoric species with excellent genome annotations - humans, mice and fruit flies (the last two also being well-established animal models for the functional study of human spermatogenesis) - and observed that, as previously reported12,22, male germ cells have a generally more diversified transcriptome than their somatic counterparts (as measured by the percentage of the entire genome that each cell type expresses). Yet, this increased complexity was mainly restricted to extremely permissive minimum average expression cut-offs ranging from >0.01 to >0.5 transcripts per million (TPM), suggestive of a significant degree of transcriptional noise or other previously hypothesized translation-independent roles for lowly expressed transcripts23. Indeed, the use of a more suitable threshold for protein-coding transcripts (TPM >1)24 revealed that, on average, metazoan male germ cells tend to express approximately 10,000 genes (9,468 ± 498 to 12,195 ± 1,144, depending on the species), corresponding to half of the entire protein-coding genome (48.0 - 67.9%; Fig. 1b and Sup. Table 1). In both humans and mice, this range is comparable to that observed in somatic cell types such as neurons and enterocytes (53.4 - 53.7% and 37.3 - 45.6% of the genome, respectively). Only in fruit flies, a species with a smaller genome (13,947 protein-coding genes vs. 22,802 and 22,287 in humans and mice, respectively), the diversity of the male germ cell transcriptome diverged substantially from that found in representative somatic tissues (with an average of 67.9 ± 3.6% of the genome expressed in germ cells vs. 33.3 ± 12.5% in the soma). It should be noted that this percentage may be higher in fruit flies due to the presence of transcripts of somatic origin in the germ cell dataset (see Sup. Table 1). Collectively, these observations suggest that the average functional requirement of a metazoan male germ cell corresponds to the expression of approximately 10,000 protein-coding genes. Such a number brings a significant but largely overlooked functional constraint to the general notion that almost all protein-coding genes are transcribed in the male germ line.

The male germ cell transcriptome has an old evolutionary origin

We next assessed the contribution of deeply conserved genes to the male germ cell transcriptome. For this we used phylostratigraphy, a technique that determines the evolutionary age of groups of homologous genes (defined as orthogroups in our analysis, including both orthologs and paralogs) by mapping their last common ancestor in a species tree25. We assembled our tree based on the proteomes of 25 phylogenetically representative eukaryotic species, and assigned each orthogroup to the oldest phylogenetic node (phylostratum) it could be traced back to (Fig. 1c and Sup. Fig. 1). This way, phylostrata ranked 1 to 5 were the oldest and contained orthogroups common to all Metazoa, while phylostratum 16 contained the youngest, species-specific orthogroups. A total of 113,757 orthogroups were identified in the 25 representative eukaryotes, 85.5% of which (97,270) were species-specific when adding all phylostrata. By using TPM >1 as minimum expression cut-off (to minimize the effects of widespread, low-level transcription in the male germ line), we observed that 48.0 - 67.9% of the entire protein-coding genome was expressed in male germ cells, depending on the species. Despite the rapid divergence typically associated with reproduction-related genes, we found that the majority (65.2 to 70.3%) of genes expressed in male germ cells mapped to the oldest-ranking phylostrata containing orthogroups common to all Metazoa, henceforth referred to as deeply conserved genes (Fig. 1d). Indeed, in all three tested species, this percentage was not significantly different from that recorded in the transcriptome of the somatic cell types (also at a threshold of TPM >1). The fraction of deeply conserved genes that were ubiquitously expressed (proxy to their involvement in cellular housekeeping processes) varied from 49.7 to 63.4% (Sup. Fig. 2a-b). This strongly suggests that almost half of all deeply conserved genes expressed by male germ cells are involved in more specific roles than just the maintenance of basal cellular functions.

By summing the products of the age of all expressed genes and their expression levels at a given developmental stage (a metric known as the transcriptome age index - TAI26) we determined the transcriptome age of the different male germ cell stages (Fig. 1e). As recently reported in mammalian spermatogenesis20, meiotic and pre-meiotic cells across all three tested species had lower TAIs than post-meiotic cells, indicative of older transcriptomes. This trend was less obvious in the fruit fly as in this species post-meiotic transcription is largely residual27. Collectively, we observed that, both in vertebrates and invertebrates, the male germ cell transcriptome has an old evolutionary origin that is tempered by the increased expression of younger genes at later developmental stages. Our data, indicative of a core spermatogenic program spanning more than 600 million years of evolution, significantly expand the breadth of previous observations in the mammalian lineage18–20, while emphasizing the potential relevance of using conserved expression between distant animal species to gain insight into the molecular basis of human infertility.

Deeply conserved genes are central components of the male germ cell transcriptome

We next addressed the possible significance of the abundant expression of deeply conserved genes in the male germ line. For this, we focused on meiosis, a fundamental process conserved through eukaryotes. Besides the existence of an ancient meiotic toolkit required for recombination and reduction division across sexually reproducing species28, the mitosis-to- meiosis transition in the male germ line is also home to a transcriptional burst associated with the acquisition of male germ cell identity29. The transcripts produced in this burst are mainly required for post-meiotic development, with their translation being shifted towards the later spermatogenic stages30. Accordingly, when analyzing the expression pattern of all human and mouse genes annotated to the “spermatid differentiation” Gene Ontology (GO) term (GO:0048515), we observed that only a small fraction of these was expressed primarily/exclusively after meiosis [5.7% (13 out of 228) and 8.7% (25 out of 289) in humans and mice, respectively]. Similar to previous reports in fruit flies27,29,31, most of the mammalian spermatid differentiation genes were found to be expressed already at the meiotic stage, with 82.9% (189 out of 228, in humans) and 77.9% (225 out of 289, in mice) being expressed in spermatocytes or in spermatocytes and spermatogonia. These results show that the meiotic transcriptome provides a particularly suitable entry-point into the genetic basis of the core spermatogenic program. Hence, we assembled protein-protein interaction (PPI) networks based on the transcriptome data of human, mouse, and fruit fly spermatocytes, where nodes represent all expressed genes and edge weights indicate the probability of the connected genes contributing to a specific function according to the STRING database32 (Fig. 1f). The structure of these networks reflects the multiple genetic interdependencies responsible for cellular function, as illustrated by the characteristic clustering of functionally related genes into topologically defined modules33. Network edges were filtered to remove low confidence interactions (those with a combined confidence score <0.5, Sup. Fig. 3a-c), and, as before, only genes expressed at TPM >1 were included to minimize the confounding effects of widespread low-level transcription. The resulting networks contained between 7,961 and 11,322 genes, depending on the species, and an average of 301,036 ± 66,416 edges.

Consistent with phylostratigraphy, the spermatocyte PPI networks included a substantial fraction of genes conserved across all Metazoa (45.1 to 58.1% of all genes; Fig. 1g). As a starting point for our analyses, we set out to determine to what extent the previously reported increased connectivity of conserved genes in PPI networks34,35 is maintained in a cell type characterized by rapid rates of evolution. For comparison purposes, we assembled, as before, PPI networks based on the transcriptome data of human, mouse, and fruit fly enterocytes (a somatic cell type with a comparably young transcriptome; Sup. Fig. 2b). These somatic cell networks, albeit smaller due to lower transcriptome diversity in enterocytes (Fig. 1b), were comparable in size to those of the spermatocytes, containing between 3,310 and 9,837 genes, depending on the species, and an average of 203,296 ± 117,545 edges (Sup. Fig. 3a-d). Similarly to what we observed in enterocytes, deeply conserved genes in the spermatocyte PPI networks were significantly more connected than their non-conserved counterparts (higher degree centrality), and their interactors were themselves more connected than those of non-conserved genes (higher page rank, Kolmogorov-Smirnov test for the two analyses; Fig. 1h; Sup. Fig. 4). Although the connectivity properties of the spermatocyte conserved genes were lower than those expressed in enterocytes, they were still salient enough for machine learning classifiers to predict reliably if a gene was conserved just based on spermatocyte network features. More specifically, using a Random Forest classifier, the AUC (area under the receiver operating characteristic curve) score was 0.74, 0.75, and 0.82 in the human, mouse, and fruit fly spermatocyte datasets, respectively (Fig. 1i). Classification performance was equally high when using a different supervised learning algorithm (linear Support Vector Machine) and an alternative performance metric (precision and recall; Sup. Fig. 5).

To address a possible ascertainment bias associated with more available information on conserved genes in the STRING database, we tested to what extent degree centrality and page rank were affected by network rewiring. In this approach, a variable percentage (20 to 100%) of all edges is randomly shuffled across the network, thus diluting any latent biases in the datasets. We observed that both network properties remained higher in conserved genes even when 80% of all edges were rewired in the spermatocyte PPI networks (Sup. Fig. 6). Such result suggests that the increased connectivity of conserved genes is mostly driven by their intrinsic properties, rather than by differences stemming from the amount of source data available. Based on these analyses, we conclude that despite the testis being a rapidly evolving organ and a preferential birthplace for new genes, deeply conserved genes remain central components in the male germ cell transcriptome of evolutionarily distant species. Accordingly, we decided to explore this increased connectivity as a means of gaining insight into the conserved molecular basis of metazoan spermatogenesis.

A core component of the ancient genetic program of spermatogenesis

We have previously shown that it is possible to simplify the complexity of biological networks by removing redundancy and only retaining the most relevant interactions that sustain the dynamics of the system36. To do so, one can extract from a network its distance backbone, i.e., the small subset of metric edges that is sufficient to compute all shortest paths in the original network37.

This technique retains the minimum number of required edges to preserve all network nodes, as well as all of the natural hierarchy of connections and community structures38. In the context of biological networks, such distance backbone uncovers the subgraph of gene interactions more likely to convey the key functional characteristics of the gene expression program39.

By extending orthology to the backbone, we have developed a new type of multilayer network: the orthoBackbone. It expands on the concept of interologs40,41, as it consists of only the edges that connect the same pair of orthologs in the network backbone of different species (Fig. 2a). Since this technique excludes all backbone edges (and corresponding nodes) that do not involve the same orthologs across multiple species, it ultimately identifies a subset of conserved genes whose functional interactions at the protein level are part of the distance backbone across evolution. In this regard, the orthoBackbone offers access to what can be considered core features of the genetic program of a given cellular state. Importantly, the use of fruit flies (an animal without male crossing over and recombination) in this inter-species approach significantly reduces the contribution of the ancient meiotic toolkit to the orthoBackbone, thus emphasizing functional interactions associated with spermatogenesis but not with meiosis per se.

Functional analysis of the conserved genetic program of male germ cells. a-

The orthoBackbone methodology. First, the most relevant associations are determined by defining the individual metric backbones (based on shortest paths) of protein-protein interaction (PPI) networks from different species. Of the backbone edges (in green), those connecting the same orthologous genes across the different species are selected as part of the evolutionarily conserved orthoBackbone (in red, with asterisks). In the case of a one-to-many conserved edge relationship, inclusion depends on at least one of the multiple edges being part of the backbone. Letters depict different genes, B’ and B’’ are paralogs, and numbers indicate distances between genes. b- The orthoBackbone represents less than 3% of all functional interactions (edges) in the spermatocyte PPI networks. c- The orthoBackbone connects >70% of all conserved genes expressed in spermatocytes. Gene conservation (across Metazoa) was defined based on eggNOG orthogroups. d- orthoBackbone genes are preferentially involved in gene expression regulation compared with other equally conserved genes. Charts represent the top 5 terms of an unfiltered gene ontology (GO) enrichment analysis for biological processes of the human male germ cell orthoBackbone. False discovery rate ≤0.05; see Sup. Fig. 7 for the expanded GO analyses. e- The male germ cell orthoBackbone reveals a core set of 79 functional interactions between 104 gene expression regulators of spermatogenesis. Solid dots indicate genes with testis specific/enriched expression. Post-transc. reg.: Post-transcriptional regulation; RNA mod.: RNA modification. f- Conserved mitosis-to-meiosis transcriptional burst genes were defined based on their upregulation at mammalian meiotic entry and/or downregulation at meiotic exit. In both cases, genes also had to be expressed in insect spermatogenesis. Green lines link orthologs (920 in fruit flies, 797 in humans and 850 in mice) based on eggNOG orthogroups. Expression level in normalized absolute log(FPKM+1). g- An in vivo RNAi screen in fruit fly testes uncovers the functional requirement of 250 conserved transcriptional burst genes (27.2%) for male reproductive fitness. Silencing of the 920 genes was induced at the mitosis-to-meiosis transition using the bam-GAL4 driver. Color-code for the recorded testicular phenotype as in “h”. Results reflect a total of four independent experiments. Threshold for impaired reproductive fitness (red horizontal line) corresponds to a 75% fertility rate (>2 standard deviations of the mean observed in negative controls). h- Conserved transcriptional burst genes are required for diverse spermatogenic processes. Testicular phenotypes of the 250 hits were defined by phase-contrast microscopy and assigned to five classes based on the earliest manifestation of the phenotype. i- Transcriptional burst genes reveal 161 new, evolutionarily conserved regulators of spermatogenesis (64.4% of all hits, homologous to 179 and 187 in humans and mice, respectively). Phenotype novelty was defined by lack of previously published evidence of a role in male fertility/spermatogenesis in humans, mice or fruit flies. j- All data acquired in this screen are freely available in the form of an open-access gene browser (Meiotic Navigator).

We observed that the orthoBackbone drastically reduced the size of the spermatocyte PPI networks, retaining only 1.7 to 2.7% of all edges, depending on the species (Fig. 2b). This almost residual fraction of edges nevertheless contained 70% of all deeply conserved genes in the network (Fig. 2c and Sup. Table 2), further illustrating that strong functional relationships between deeply conserved genes are also maintained in the male germ line. GO term enrichment analysis for biological processes revealed that orthoBackbone genes were preferentially involved in gene expression and protein regulation, in contrast with the other (not retained) conserved genes that were mainly associated with cell signaling pathways (Fig. 2d and Sup. Fig. 7). To better understand to what extent the orthoBackbone can provide insight into cell-type specific processes, we determined the degree of similarity between the spermatocyte and enterocyte orthoBackbones. We noted a modest overlap between both subnetworks, with just 16.2% to 18.0% of common edges depending on the species (against background expectations of 33.6 to 61.5%, based on Jaccard similarity index scores between the PPI networks of both cell types; Sup. Fig. 3d).

Using GO annotations, we further selected spermatocyte orthoBackbone interactions involving gene expression regulators previously linked to spermatogenesis. By doing so, we obtained a subset of 79 functional interactions between 104 human genes (out of the total of 3,596 in the orthoBackbone; Fig. 2e). These were ascribed to a wide gamut of regulatory processes and involved noteworthy spermatogenesis genes such as RFX2 (transcription), CDYL (chromatin remodeling) and BOLL (translation), among others42–44. Of the 104 genes, 34 were testis- specific/enriched while 60 had low tissue specificity, consistent with previous observations that cell identity relies on the integration of cell type-specific genes with more broadly expressed regulators45. Based on its conservation in the orthoBackbone of evolutionarily distant species and on the significant spermatogenic role of the intervening genes, we posit that this subnetwork of 79 functional interactions involving 104 genes likely represents a core component of the conserved genetic program of spermatogenesis. This number fits well with previous estimates, based on expression trajectories of 1:1 orthologous genes, of an ancestral program of amniote spermatogenesis constituted by 389 genes20.

Male germ cell identity has a broad functional basis

We next set out to determine the functional consequences of disrupting the spermatocyte orthoBackbone. Since the latter represents a still sizeable interaction subnetwork consisting of more than 3,000 genes, we decided to select candidate genes likely to affect the acquisition and maintenance of male germ cell identity. For this, we again took into account how the mitosis-to- meiosis transcriptional burst activates the expression of spermatogenesis genes - a process that has been likened to that of cellular reprogramming46,47. We reasoned that genes that are upregulated at meiotic entry and/or downregulated at meiotic exit (thus part of the transcriptional burst) likely represent instructive elements for the male germ cell state, hence preferential routes to tamper with this cellular identity.

Through differential gene expression analysis, we identified, in human and mouse spermatocytes, all expressed genes that shared a similar upregulation at meiotic entry and/or downregulation at meiotic exit. Of these 970 mammalian differentially expressed genes (DEGs), we selected as candidates for functional assessment only those that were also expressed in fruit fly germ cells (Fig. 2f). We did not take into account differential expression in the latter species, since the largely residual levels of post-meiotic transcription in fruit fly spermatogenesis thwarts direct comparisons with the mammalian system27. The resulting 920 fruit fly genes (homologous to 797 and 850 in humans and mice, respectively) were silenced specifically at the mitosis-to-meiosis transition by Drosophila in vivo RNAi, using the well-established bam-GAL4 driver48. Of these, a total of 250 (27.2%) were essential for male fertility, as their silencing resulted, upon mating with wildtype females, in egg hatching rates below the cut-off of 75% (>2 standard deviations of the mean observed in negative controls: 91.6 ± 8.5%; Fig. 2g). We observed that 76.0% of the hits (190 out of 250) were part of the orthoBackbone, compared with 60.0% (402 out of 670) in genes without any obvious effect on male reproductive fitness upon silencing, with this difference being statistically significant (two-sided Fisher’s exact test P<0.0001). Cytological analysis of all 250 genes by testicular phase-contrast microscopy in the RNAi males revealed diverse origins for the infertility phenotype, with the earliest manifestation of cellular defects ranging from the pre-meiotic to the mature gamete stage (Fig. 2h). By exploring publicly available information (see Methods), we determined that 161 (64.4%) of all hits had never been previously associated with male reproduction in any species (Fig. 2i). Accordingly, these 161 new deeply conserved spermatogenesis genes (homologous to 179 and 187 in humans and mice, respectively) represent both a significant advance in our understanding of the genetic basis of male germ cell development, and a valuable resource to explore from a precision medicine perspective. To facilitate open access to this information, we made all data available in the form of a user-friendly gene browser (https://pages.igc.pt/meionav; Fig. 2j). Overall, by merging our results with previously published data we conclude that at least 43.5% of the mitosis-to-meiosis transcriptional burst (our 250 hits + 150 previously reported male fertility genes that gave a negative result in our assay) is required for a diverse range of germ cell functions. This illustrates the pervasive influence of the mitosis-to-meiosis transition on multiple facets of the spermatogenic program and argues for a broad functional basis underlying male germ cell identity.

An ancient regulator of human spermatogenesis

One of our 161 newly identified spermatogenesis genes - the Drosophila RING finger protein 113 (dRNF113, a spliceosomal component also known as mdlc49) - emerged as a particularly interesting germ cell regulator from a clinical perspective. More specifically, by analyzing our in-house whole exome database containing sequencing data from 74 cases of human male meiotic arrest, we identified an infertile man harboring a homozygous loss of function (LoF) variant in a human paralog (RNF113B). This frameshift variant c.556_565del;p.(Thr186GlyfsTer119) leads to the abrogation of the protein’s two functional domains (a C3H1-type zinc finger and a RING finger) and to its truncation (Fig. 3a and Sup. Fig. 8a). The identified man (M1911) is part of a consanguineous family of Middle-Eastern ancestry and shared the homozygous state of the RNF113B LoF variant with his equally infertile brother, but not with his fertile heterozygous brother (Sup. Fig. 9a-c and Sup. Information). We excluded the possibility that this genetic variant is an indicator of another, nearby variant that is homozygous in the affected individuals and represents the actual cause of infertility in both men (Sup. Information). Remarkably, the results from the testicular biopsy of M1911 revealed an equivalent meiotic arrest phenotype to that observed in dRNF113-silenced fruit flies. Indeed, spermatocytes were, by far, the predominant cell type in the male gonads of both species: they were the most advanced cell stage observed in 89.0% of all assessed human seminiferous tubules (vs. 9% in controls) and occupied an average of 64.1% of the entire testis area in fruit flies (vs. 12.3% in controls; Fig. 3b-c and Sup. Fig. 8b-c). Early (round) spermatids were practically absent in both species, with cellular debris accumulating in the post-meiotic region of the fruit fly gonad. The similar reproductive impairment phenotypes recorded in both species highlights the often-complex relationship between tissue specificity and functional significance. In insects, dRNF113 is a broadly expressed gene essential for organismal viability49, whose particular spermatogenic function was uncovered by germ cell-specific silencing. Both roles were functionally uncoupled in primates via the existence of two RNF113 copies, with the correctly spliced form of RNF113B being mainly testis-specific50. The existence of an independent duplication event in the rodent lineage offered us the opportunity to further explore this concept. Compared with its human ortholog (RNF113B), the mouse duplication (Rnf113a2) is broadly expressed, similarly to Rnf113a1, human RNF113A and fruit fly dRNF113. Yet, the generation of a whole-body Rnf113a2 knockout (KO) mouse revealed a comparable phenotype to that of the human RNF113B LoF variant: severe spermatogenic impairment without any overt somatic defects. When compared to wildtype littermates, male homozygous Rnf113a2KO/KO mice were sterile and had visibly smaller testes that were largely devoid of germ cells (SCO: Sertoli cell-only phenotype) except for the rare presence of spermatogonia and spermatocytes (Fig. 3d and Sup. Fig. 8d-f).

Deeply conserved regulators of human spermatogenesis. a-

Similar domain structure of the RNF113 proteins in humans (RNF113B), fruit flies (dRNF113) and mice (Rnf113a2). All contain a C3H1-type zinc finger and a RING finger domain. Numbers indicate amino acid residue position. b- Human RNF113B is required for meiotic progression past the primary spermatocyte stage. Testicular histology of M1911 [RNF113B loss of function (LoF) variant] and of a control sample with normal spermatogenesis. See Sup. Fig. 8b for phenotype quantification. Arrowheads: primary spermatocytes; arrows: spermatids. Scale bars: 100 μm (overview), 50 μm (insets), and 10 μm (meiotic region). c- Silencing fruit fly dRNF113 results in meiotic arrest. Phase-contrast microscopy. See Sup. Fig. 8c for phenotype quantification. Arrowheads: primary spermatocytes; asterisks: early (round) spermatids; arrows: late (elongating) spermatids; sv- seminal vesicle. Scale bars: 50 μm (whole testis) and 20 μm (meiotic region). d- Mouse Rnf113a2 is essential for spermatogenesis. Testes of whole-body homozygous Rnf113a2 knockout mice (Rnf113a2KO/KO) are essentially devoid of germ cells, with the rare occurrence of meiotic and pre-meiotic stages. The testicular histology of a wildtype littermate control (Rnf113a2WT/WT) is presented for comparison (normal spermatogenesis). See Sup. Fig. 8d-g for additional data. Arrowheads: primary spermatocytes; arrows: spermatids. Scale bars: 50 μm (overview) and 20 μm (intratubular region). e- Human RNF113B is predominantly expressed at meiotic entry. Data analyzed from our recently published single cell RNA-Seq atlas of normal spermatogenesis51. Usg.: undifferentiated spermatogonia; Dsg.: differentiated spermatogonia / preleptotene; Lp.: leptotene; Zg.: zygotene; Pc.: pachytene; Dp.: diplotene; M: meiotic divisions; Rsp.: round spermatids; Esp.: elongating spermatids. f- The nuclear levels of the fruit fly dRNF113 protein increase at meiotic entry. Images are maximum projections of the entire nuclear volume. Spermatocytes correspond to late prophase I cells. Dotted lines delimit the nuclear envelope (as assessed by fluorescent wheat germ agglutinin). Scale bar: 5 μm. a.u- arbitrary units. ****p <0.0001 (unpaired t-test). g- RNF113B is required for normal gene expression during human spermatogenesis. Differential gene expression (DGE) analysis of RNA-Seq data obtained from testicular biopsies of M1911 (RNF113B LoF variant, left and right testis) and of three controls with normal spermatogenesis. Down and upregulated genes in blue and red, respectively. orthoBackbone differentially expressed genes (DEGs) are outlined. FC: fold change. FDR: false discovery rate. Edge disruption corresponds to the number of orthoBackbone edges containing at least one DEG. h- dRNF113 regulates gene expression in the fruit fly male gonad. Whole testes samples (in triplicate) in both experimental conditions. i- Network of functional associations between orthoBackbone genes downregulated both in the RNF113B LoF and dRNF113 RNAi. Node size indicates the result of the page rank metric in the spermatocyte protein-protein interaction network (measure of the connectivity of interacting genes), and color specifies if the gene has a known role in spermatogenesis (in any species). Testicular phenotype of men affected by variants in HSPA2 and KPNA2 (red nodes) are depicted in “j”. Edge thickness indicates STRING functional association scores, and color specifies the main source of data for the associations. j- LoF variants in the orthoBackbone genes HSPA2 and KPNA2 are potentially associated with human male infertility. Testicular histology of individuals M2190 and M2098 (HSPA2 and KPNA2 variants, respectively) reveals a complete loss of germ cells (Sertoli cell- only phenotype). Arrowheads: Sertoli cells. Scale bars: 100 μm (overview), 50 μm (insets), and 10 μm (intratubular region).

Strengthening our interest in the RNF113 genes was their possible involvement with male germ cell identity. By analyzing our recently published single-cell RNA-Seq dataset of normal human spermatogenesis51, we observed that RNF113B was predominantly expressed at the mitosis-to- meiosis transcriptional burst, peaking at the diplotene stage of prophase I (Fig. 3e). This mirrored the protein localization pattern of fruit fly dRNF113, characterized by a substantial nuclear accumulation in primary spermatocytes (Fig. 3f). Also in mice, Rnf113a2 was expressed at the transcriptional burst, with an earlier expression in undifferentiated spermatogonia being the likely cause of the slightly more severe histological phenotype of the Rnf113a2KO/KO mice (Sup. Fig. 8g). Consistent with their role in gene expression, the analysis of M1911’s testicular transcriptome via RNA-Seq revealed the deregulation of 21.7% of all expressed genes (Fig. 3g). Notably, the deregulation detected in M1911’s testicular transcriptome had a clear impact on the spermatocyte orthoBackbone, with 30.0% of its network edges being disrupted (i.e., containing at least one deregulated gene). A similar effect was patent in the dRNF113 RNAi, with 20.3% of the fruit fly testicular transcriptome being deregulated, resulting in the disruption of 26.3% of the orthoBackbone (Fig. 3h). For comparison, the silencing of Prp19, another transcriptional burst gene and spliceosomal component that RNF113 proteins associate to52, had a lower effect on the orthoBackbone with 15.2% of disrupted edges (significantly lower than in the dRNF113 RNAi; two-sided Fisher’s exact test P<0.0001), despite a similar meiotic arrest phenotype (Sup. Fig. 10a-d). Based on the above, we conclude that the RNF113 genes have retained a key role in spermatogenesis for more than 600 million years of evolution. Furthermore, they provided us the means to interfere with a sizable fraction of the male germ cell orthoBackbone in humans and fruit flies - an advantage we next explored from a clinical perspective.

The orthoBackbone as an ancillary tool in clinical genetics

Male infertility is a complex human disease with a poorly defined genetic component. This contributes to a low causative diagnostic yield (typically just 30%), and to a paucity of clinically validated genes [currently just 104, in contrast with the more than 700 already associated with other clinical disorders such as intellectual disability53,54]. Since male infertility affects up to 12% of all men55, addressing such knowledge gap is an issue of clear medical importance. To narrow the gap, we explored the possibility that spermatogenesis is particularly sensitive to disturbances in the male germ cell orthoBackbone. Thus, we harnessed the sizeable effect of RNF113B on this network as a means of identifying additional genetic causes of human infertility. Indeed, we noted that 38.0% (30 out of 79) of the functional interactions that define our previously identified core component of the conserved genetic program of spermatogenesis were disrupted in M1911 (i.e., these interactions contained at least one differentially expressed gene; Sup. Fig. 11). In particular, we observed that 19.2% (20 out of 104) of this gene subset was found to be downregulated, compared with 10.0% (1,685 out of 16,824) of downregulated genes in the testicular transcriptome of M1911, with this difference being statistically significant (two-sided Fisher’s exact test P=0.0046; Fig. 3g). Therefore, we posited that the similar spermatogenic impairment recorded in the human RNF113B LoF variant and on the fruit fly dRNF113 RNAi ultimately reflected the downregulation, in both species, of a common set of orthoBackbone genes.

By defining the overlap between differentially expressed orthoBackbone genes in the testicular transcriptome of M1911 and of the dRNF113 RNAi, we identified 61 conserved human genes that were similarly downregulated in both species. These formed a connected network (based on STRING data) suggestive of their functional co-involvement in related biological processes (Fig. 3i). Of the 61 genes, 27 had already been linked to male germ cell defects in different species: four were clinically validated male infertility genes (CDC14A, CFAP91, DNAI1 and DNAI2), and the other 23 had been previously reported in animal models (see Methods). The fact that among the latter was BOLL, one of the oldest known metazoan gametogenesis genes44, was a particularly noteworthy observation. We next tested if these 27 genes could be used to identify new genetic causes of human infertility. For this, we analyzed whole-exome sequencing data of 1,021 azoospermic men from the MERGE cohort56. Filtering these exomes for LoF variants in the 27 selected orthoBackbone genes revealed two new human male infertility candidate genes: HSPA2 and KPNA2.

HSPA2 - heat shock protein family A member 2 - is a molecular chaperone of the highly conserved 70-kDa heat shock protein (HSP70) gene family. HSP70 members are involved in cellular proteostasis from bacteria to human, with the mouse and fruit fly HSPA2 homologs (Hspa2 and Hsc70-1 to Hsc70-5, respectively) being required for meiotic progression past the primary spermatocyte stage57,58. Indeed, we observed that the silencing of Hsc70-1 in the fruit fly testis, also with the bam-GAL4 driver, resulted in an equivalent meiotic arrest to that of dRNF113, strongly suggesting that the misexpression of this chaperone is a central element of the latter phenotype (Sup. Fig. 12a-b). Two different heterozygous LoF variants were detected in our male infertility cohort, with this gene having a predicted autosomal-dominant inheritance (Sup. Fig. 12c-f, Sup. Information and Methods). The first variant, detected in individual M1678, is the early stop-gain variant c.175C>T;p.(Gln59Ter) that truncates more than 90% of the protein, likely leading to nonsense-mediated decay. There is only one individual listed in the gnomAD database with this variant (minor allele frequency = 0.0004%). The second, identified in individual M2190, is the frameshift variant c.1806dup;p.(Glu603ArgfsTer81) which affects the distal end of the protein’s nucleotide-binding domain. This variant is not listed in the gnomAD database. Histopathological analysis of M2190’s testicular tissue revealed a SCO phenotype in 259 assessed seminiferous tubules, a likely aggravation of the extensive cell death accompanied by tubule vacuolization reported in Hspa2 mutant mice59 (Fig. 3j).

KPNA2 - karyopherin subunit alpha 2 - is a nuclear importin that functions as an adapter protein in nucleocytoplasmic transport. Its mouse ortholog (Kpna2) is required for the nuclear accumulation of Hop260, a conserved regulator of meiotic progression from yeast to mammals61,62. The main function of Hop2 is to repair meiotic DNA double-strand breaks (DSBs), with the corresponding mouse mutant being characterized by a primary spermatocyte arrest coupled to extensive cell death. Since male fruit flies dispense with meiotic DSBs63, the inclusion of KPNA2 in the male germ cell orthoBackbone seems counter-intuitive. Yet, Drosophila importin alpha 2 (dKPNA2, also known as Pendulin) is required for post-meiotic development64, with its silencing specifically at meiotic entry being associated with aborted spermiogenesis (Sup. Fig. 13a-b). This post-meiotic function might also be present in its mammalian orthologs, as suggested by the nuclear localization of Kpna2 in elongating mouse spermatids60. We detected two heterozygous KPNA2 LoF variants in our cohort of infertile men, with this gene also having a predicted autosomal-dominant inheritance (Sup. Fig. 13c-g, Sup. Information and Methods). The splice- site variant (c.667-2A>G;p.?) in individual M1645 is predicted to disrupt the correct splicing of intron 6 due to the loss of an acceptor site. Segregation analysis revealed the de novo occurrence of this variant, strongly supporting both its pathogenicity and our recent results, pointing to the important contribution of heterozygous de novo mutations to male infertility65,66. The other KPNA2 LoF variant, in individual M2098, is the frameshift variant c.1158_1161del;p.(Ser387ArgfsTer14) which affects 27% of the protein, including its minor nuclear localization signal binding site. Both KPNA2 variants are not listed in the gnomAD database. The available testicular histopathology report for the latter individual lists a SCO phenotype as the cause of the azoospermia, again suggestive of the possible deterioration of an initially meiotic phenotype (Fig. 3j).

Based on three in silico prediction tools (Sup. Information and Methods), both HSPA2 and KPNA2 can represent haploinsufficient genes and/or have autosomal dominant inheritance, with the evidence being stronger for KPNA2. Together with the identified de novo occurrence of one of its LoF variants, this makes KPNA2 a strong novel autosomal-dominant candidate gene while more data is still needed before drawing similar conclusions on HSPA2.

In summary, by experimentally disrupting the male germ cell orthoBackbone across evolutionarily distant species, we were able to uncover two new candidate genes for human infertility (HSPA2 and KPNA2) in addition to RNF113B, affecting a combined total of 5 individuals. This successful merger between basic and clinical research highlights the advantages of comparative biology as means of dampening inter-species differences in reproductive physiology, while providing a conceptual framework for a more efficient prioritization of clinically-relevant genetic variants in human reproductive disease.

Discussion

In summary, we have identified a conserved male germ cell gene expression program, spanning more than 600 million years of evolution, that can provide clinical insight into the molecular basis of human infertility. We estimate the average functional requirement of a metazoan male germ cell to correspond to the expression of approximately 10,000 protein-coding genes, the majority of which are deeply conserved. Indeed, despite the testis being a rapidly evolving organ and frequent birthplace of genetic novelty, deeply conserved genes are central components of the male germ cell transcriptome, with younger genes being preferentially expressed in late spermatogenesis. We further pinpoint 79 functional interactions between 104 transcriptional regulators that, owing to their salient network properties across different species, likely define a core component of the ancient genetic program of spermatogenesis. By functionally interfering with 920 deeply conserved genes associated with male germ cell identity, we uncover 161 previously unknown spermatogenesis genes required for a broad range of germ cell functions, as well as three new potential genetic causes of human infertility.

We propose that the transcriptional identity of metazoan male germ cells is built around a relatively small network of deeply conserved gene interactions with an overarching functional impact. These observations may provide a unifying genetic basis for the deep conservation of fundamental germ cell biological processes such as sperm motility and gamete fusion67,68. The existence of a shared genetic identity in metazoan spermatocytes can be regarded as a ramification of an ancestral multipotency program already present in germ line precursor cells17. Indeed, even in species with divergent germ line segregation strategies, conserved functional interactions at the post-translational level are required for primordial germ cell specification69,70. It should be noted that our conclusions were based on a small fraction of the diversity of all metazoan species, as a result of data availability limitations and our focus on well-characterized gonochoric animal models of human spermatogenic impairment. The advent of more comprehensive gene expression datasets from less-studied organisms will surely provide a more encompassing context to cross-species comparisons of this nature.

The actual benefit of comparative biology for the identification of new genetic causes of human disease is often a contentious topic71. Central to this debate is the moderate success in translating animal data to the clinical setting, coupled with the fact that the vast majority of human genetic variants are not shared with other species72. By focusing on the deep evolutionary past of human spermatogenesis, we have identified 79 functional interactions that likely represent a core component of the conserved genetic program of male germ cells, and 179 novel functionally validated candidate genes. Exploring to what extent these core functional interactions reflect direct protein-protein contacts is an important area for future research. In this regard, the recent development of robust in silico predictive systems, such as AlphaFold-Multimer, will greatly facilitate the implementation of such studies73.

The translational application of our cross-species platform in human reproductive healthcare has so far uncovered three new genes associated with human male infertility (RNF113B, HSPA2 and KPNA2). The assumed autosomal dominant inheritance of HSPA2 and KPNA2 awaits further confirmation by additional data from independent cohorts and segregation analyses, but stands in line with other well-established male infertility genes such as DMRT1, NR5A1 and SYCP253. Also, this information can potentially be harnessed in other clinically relevant contexts: the spermatocyte orthoBackbone can be further explored to pinpoint key pathways that may facilitate the reconstitution of meiotic entry in vitro, as well as to maximize the effectiveness of exome sequencing approaches in infertile men through the inclusion of the orthoBackbone genes in prioritization lists for disease-causing variants. Conversely, future iterations of our analytical pipeline can be specifically tailored for the identification of newly emerged genes that may have acquired key roles in reproductive processes, such as those involved in species-specific gamete recognition74.

Taken together, our results emphasize the often-overlooked contribution of evolutionary history to human disease and illustrate how interdisciplinary research can significantly expand our knowledge of fundamental cellular processes. Accordingly, all the code required for repurposing our analytical pipeline to other cell types and pathologies is available as an open-access resource at https://github.com/rionbr/meionav. These resources will likely contribute to a renewed appreciation of comparative biology in the medical field.